010CreateTestData
=================
Purpose
-------
* Download files from [FDA.gov](http://www.fda.gov/Drugs/InformationOnDrugs/ucm142438.htm) and create R data frames
* Join the Product dataset to the Package dataset
* Classify the Package Codes into format types
  * 4-4-2
  * 5-3-2
  * 5-4-1
* Sample 100 entires from each Package Code format type
* Export to tab-delimited text files


Download
--------
Download the zip archive from the [FDA.gov](http://www.fda.gov/Drugs/InformationOnDrugs/ucm142438.htm) website. Unzip the archive. Read the two datasets into data frames, `dProd` for the Product Code table and `dPack` for the Package Code table.

Code is from [Stackoverflow](http://stackoverflow.com/a/3053883/1427069).
```{r}
url <- "http://www.fda.gov/downloads/Drugs/DevelopmentApprovalProcess/UCM070838.zip"
f <- tempfile()
download.file(url, f)
t <- unz(f, "product.txt")
dProd <- read.delim(t)
t <- unz(f, "package.txt")
dPack <- read.delim(t)
unlink(f)
```

Rename variables to lower case.
```{r}
names(dProd) <- tolower(names(dProd))
names(dPack) <- tolower(names(dPack))
```


Join
----
Join the Product dataset to the Package dataset.
```{r}
require(dplyr, quietly=TRUE)
d <- inner_join(dProd, dPack)
```


Classify
--------
Classify the Package Codes into format categories
* 4-4-2
* 5-3-2
* 5-4-1
```{r}
isFmt442 <- grepl("[[:alnum:]]{4}-[[:alnum:]]{4}-[[:alnum:]]{2}", d$ndcpackagecode)
isFmt532 <- grepl("[[:alnum:]]{5}-[[:alnum:]]{3}-[[:alnum:]]{2}", d$ndcpackagecode)
isFmt541 <- grepl("[[:alnum:]]{5}-[[:alnum:]]{4}-[[:alnum:]]{1}", d$ndcpackagecode)
fmt <- rep(NA, length(dPack$ndcpackagecode))
fmt[isFmt442] <- "4-4-2 format"
fmt[isFmt532] <- "5-3-2 format"
fmt[isFmt541] <- "5-4-1 format"
table(fmt, useNA="ifany")
d$fmtPackageCode <- format(fmt)
```


Sample
------
Use the `ndcpackagecode` as the sampling frame. Sample 100 entires from each Package Code format type.
```{r}
frameSampling <- list(
  type442=d$ndcpackagecode[d$fmtPackageCode == "4-4-2 format"],
  type532=d$ndcpackagecode[d$fmtPackageCode == "5-3-2 format"],
  type541=d$ndcpackagecode[d$fmtPackageCode == "5-4-1 format"])
f <- function(x) {sample(x, 100)}
s <- sapply(frameSampling, FUN=f)
```
Rearrange the sample matrix to a vector so it can be used as a lookup.
```{r}
s <- as.vector(s)
```
Create a data frame of the sampled data.
```{r}
dSample <- d[d$ndcpackagecode %in% s, ]
table(dSample$fmtPackageCode)
```


Export
------
Export to tab-delimited text files.
```{r}
f <- "dataForTesting.txt"
write.table(dSample, f, quote=FALSE, sep="\t", row.names=FALSE)
```
Print a list of variables in the dataset.
```{r}
names(dSample)
```
Print a few records from the dataset.
```{r}
head(dSample)
```
