---
title: "NPPES"
author: "Benjamin Chan"
date: "April 26, 2016"
output: 
  html_document: 
    keep_md: yes
---

R Markdown script is in [import.Rmd](import.Rmd).

National Plan and Provider Enumeration System (NPPES);
a.k.a. NPI Downloadable File, or full replacement Monthly NPI File,
National Provider Identifier Standard (NPI).

[Source](https://www.cms.gov/Regulations-and-Guidance/HIPAA-Administrative-Simplification/NationalProvIdentStand/DataDissemination.html)

Grab the current Full Replacement Monthly NPI File from
[http://download.cms.gov/nppes/NPI_Files.html](http://download.cms.gov/nppes/NPI_Files.html)

```{r}
url <- "http://download.cms.gov/nppes/NPPES_Data_Dissemination_Apr_2016.zip" # Update this URL as needed
```

Download and unzip the file

```{r}
f <- tempfile()
download.file(url, f, mode="wb")
path <- tempdir()
unzip(f, exdir = path)
unzip(f, list=TRUE)
```

Read the data file

```{r}
library(data.table)
filenames <- unzip(f, list = TRUE)$Name
hdrFile <- grep("FileHeader\\.csv$", filenames, value = TRUE)
hdrFile <- file.path(path, hdrFile)
datFile <- grep("[0-9]\\.csv$", filenames, value = TRUE)
datFile <- file.path(path, datFile)
D <- fread(datFile)
```

Substitute spaces in column names

```{r}
oldvar <- names(D)
newvar <- gsub("\\s", "_", oldvar)
setnames(D, oldvar, newvar)
```

Show some examples

```{r}
D[grep("OHSU", Provider_Other_Organization_Name), .(NPI, Provider_Other_Organization_Name)]
```
